---
title: "Legitimacy costs in a 'rigged' election: Observational and survey-experimental evidence from the 2020 US election"
author: "Cole J. Harvey"
date: "10/1/2021"
output: beamer_presentation
bibliography: C:/Users/colej/Documents/Research projects/2020_election_survey/bib_ch.bib

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(echo = FALSE)

library(tidyverse)
library(lme4)
library(kableExtra)
library(modelsummary)
library(interactions)


anes_data <- read.csv(here::here("Data", "anes_modified_long.csv"))

survey.data <- read.csv(here::here("Data", "survey data.csv"))
survey.data <- survey.data %>% mutate(biden.winner = ifelse(vignette_winning_candidate == "biden", 1, 0))
survey.data <- survey.data %>% mutate(anger.index = (anger + outrage)/2)


```

## Motivation
- Common assumption in studies of election manipulation: manipulation leads to a loss of support for the regime or system
  - Legitimacy costs [@birch2011]
  - Connected to the assumption that protest risk is major deterrent to manipulation  [@magaloni2010;@Little2012;@Fearon2011;@Hyde2014a].
- Important to our theories of manipulation, and thus to democratization and authoritarianism

## Motivation
- But:
  - Usually an *assumption*
  - Mechanism is unclear
    - Intensify anger among losers?
    - Broaden the opposition coalition?
    
## Approach
- Conceptualize legitimacy costs as feelings of anger
- 2020 election known in advance to face allegations of manipulation
  - Survey experiment presenting possible outcomes
  - ANES panel survey
  
## Findings
- Survey experiment shows both mechanisms are plausible
- ANES panel shows that, in this context, base mobilization is the main channel

## Overview of literature: election manipulation
- Election manipulation has lots of benefits [@Simpser2011]
- What holds ruling parties back? Legitimacy loss [@magaloni2010;@Little2015;@Chernykh2015]
  - Part of the story in analysis of mass electoral protest [@tucker2007]
  - However, electoral protest is relatively rare [@kalandadze2009electoral]
- Birch [-@birch2011]: "...the proportion of the population that can be expected to acquiesce to having their contribution to public choice taken from them or distorted..."
- Norris et al [-@Norris2015: Loss of trust

## Theory
- Social psychologists' theories of legitimacy rest on procedural justice [@Tyler2006]
  - individuals' beliefs that officials act following rules in an even-handed, predictable way 
- Survey research finds positive relationship between perceptions of procedural justice and legitimacy judgments, and with compliance with the law [@Levi2009;@Tyler2002;@Sunshine2003;@Reisig2014;@Hartner2008]

## Theory
- Violations of procedural justice increase feelings of anger [@Weiss1999;@Krehbiel2000;@Murphy2008;@Murphy-Berman1999].
  - Even if unfairness experienced vicariously [@Gordijn2006]
  - Effect can be contagious:  witnessing others' anger can negatively influence one's own estimation of the justice [@Hillebrandt2017].
- Finally, anger contributes to increases in political engagement and protest [@van_zomeren_put_2004;@VanStekelenburg2013a;@Jasper2014].

## Theory: base mobilization
- Anger after a flawed election mediated by partisanship and status as an electoral winner or loser [@Valentino2017;@Valentino2009]

## Theory: Coalition growth

- Research on the color revolutions and beyond suggests the opposition unity before and after the election is central to the success of post-election protests [@Bunce2006a;@Bunce2006;@Vladisavljevic2016]. 
- Tucker [-@tucker2007] argues that election manipulation serves as a focal point for citizens' grievances *generally*, not only the specific grievance of having one's preferred candidate defeated. (See also [-@Brancati2014a])
- Russia 2011: support for the protest movement did not vary significantly across the left-right economic dimension, or among nationalists and non-nationlists [@Chaisty2013].

## Hypotheses 

> Hypothesis 0: Allegations of election manipulation will lead to greater interest in collective action, mediated by feelings of anger, compared to a clean election.
> Hypothesis 1 (Base mobilization): Among those who believe an election was unfair, stronger support for the losing candidate will be associated with stronger feelings of anger.

> Hypothesis 2 (Base mobilization): Among those who are supporters of the losing candidate, increasing belief in election manipulation will be associated with stronger feelings of anger.

## Hypotheses

> Hypothesis 3 (Cross-partisan coalition): Among supporters of the winning candidate, increasing belief in election manipulation will be associated with stronger feelings of anger.

> Hypothesis 4 (Cross-partisan coalition): Among supporters of the winning candidate who believe the election was unfair, more positive ratings of the losing candidate will be associated with greater anger.

## Research design: Survey experiment

- Conducted in mid-October 2020
- 1049 respondents recruited via the research platform Prolific
- 2 x 2 vignette experiment:
  - Either Donald Trump or Joe Biden would be elected
  - Victory ascribed to falsification of the results by government officials, or without attribution
  
## Research design: ANES panel
- Two-wave panel survey (pre- and post-election) conducted by NORC, US citizens age 18 and older. 
  - 5,227 respondents
- How respondents' self-reported emotions and attitudes toward election integrity changed over time. 

```{r experiment-anger-model}

model.anger.falsify <- lm(anger.index ~ trump_rating*vignette_falsify  + ideology, data = survey.data %>% filter(vignette_laws == 0 & biden.winner == 1))
#summary(model.anger.falsify)

p_exp_anger <- sjPlot::plot_model(model.anger.falsify, type = "pred", terms = c("trump_rating [0, 9]", "vignette_falsify"), title = "Predicted values of anger index", axis.title = c("Trump rating", "Anger index"),
                                  legend.title = "Fraud treatment")

```

```{r experiment-anger-model2}

model.anger.falsify <- lm(anger.index ~ trump_rating*vignette_falsify  + ideology, data = survey.data %>% filter(vignette_laws == 0 & biden.winner == 0))
#summary(model.anger.falsify)

p_exp_anger2 <- sjPlot::plot_model(model.anger.falsify, type = "pred", terms = c("trump_rating [0, 9]", "vignette_falsify"), title = "Predicted values of anger index", axis.title = c("Trump rating", "Anger index"),
                                  legend.title = "Fraud treatment")

```

## Results: Survey experiment

```{r, echo = FALSE, fig.cap="Biden winner condition", out.width="80%"}
p_exp_anger + theme_bw()
```

## Results: Survey experiment

```{r, echo = FALSE, fig.cap="Trump winner condition", out.width="80%"}
p_exp_anger2 + theme_bw()
```


```{r anger_models_trump}
anes_data <- anes_data %>% mutate(period = factor(period))

model_ft_ei_mlm <- lmer(anger_index ~ period*fttrump*ei_index  + biden_voter +  pid7x +
                     factor(profile_gender) + factor(profile_racethnicity) + 
                     profile_educ5 +
                     profile_income + profile_newskeepup + covid_fin + covid_worry + persfin + (1 | caseid)
                    , data = anes_data)

  #summary(model_mlm)
p_ft_ei_mlm <- sjPlot::plot_model(model_ft_ei_mlm, type = "pred", terms = c("period", "ei_index [1, 3, 5]", "fttrump [0, 40, 100]"), title = "Predicted values of anger index", axis.title = c("Post-election period", "Anger index"),
                                  legend.title = "Elec. confidence")


model_ft_ei_ols <- lm(change_anger_index ~ change_ei_conf*fttrump  +  pid7x + biden_voter +
                     factor(profile_gender) + factor(profile_racethnicity) + 
                     profile_educ5 +
                     profile_income + profile_newskeepup + covid_fin + covid_worry + persfin
                    , data = anes_data %>% filter(period == 1))
  #summary(model_fttrump)
p_ft_ei_ols <- sjPlot::plot_model(model_ft_ei_ols, type = "pred", terms = c("change_ei_conf", "fttrump [0, 100]"), ci.lvl = .95, title = "Predicted change in anger index", axis.title = c("Change in elec. confidence", "Change in anger index"), legend.title = "Trump rating")


```

```{r anger_models_biden}
model_biden_ei_mlm <- lmer(anger_index ~ period*biden_voter*ei_index  + fttrump +  pid7x +
                     factor(profile_gender) + factor(profile_racethnicity) + 
                     profile_educ5 +
                     profile_income + profile_newskeepup + covid_fin + covid_worry + persfin + (1 | caseid)
                    , data = anes_data)

  #summary(model_mlm)
p_bv_ei_mlm <- sjPlot::plot_model(model_biden_ei_mlm, type = "pred", terms = c("period", "ei_index [1, 3, 5]", "biden_voter"), title = "Predicted values of anger index", axis.title = c("Post-election period", "Anger index"),
                                  legend.title = "Elec. confidence")


model_biden_ei_ols <- lm(change_anger_index ~ change_ei_conf*biden_voter  +  pid7x + fttrump +
                     factor(profile_gender) + factor(profile_racethnicity) + 
                     profile_educ5 +
                     profile_income + profile_newskeepup + covid_fin + covid_worry + persfin
                    , data = anes_data %>% filter(period == 1))
  #summary(model_fttrump)
p_biden_ei_ols <- sjPlot::plot_model(model_ft_ei_ols, type = "pred", terms = c("change_ei_conf", "biden_voter"), ci.lvl = .95, title = "Predicted change in anger index", axis.title = c("Change in elec. confidence", "Change in anger index"), legend.title = "Biden voter")

```

```{r anger_model_3way}

model_biden_ei_ft_ols <- lm(change_anger_index ~ change_ei_conf*biden_voter*fttrump  +  pid7x  +
                     factor(profile_gender) + factor(profile_racethnicity) + 
                     profile_educ5 +
                     profile_income + profile_newskeepup + covid_fin + covid_worry + persfin
                    , data = anes_data %>% filter(period == 1))
  #summary(model_fttrump)
p_bv_ft_ei_ols <- sjPlot::plot_model(model_biden_ei_ft_ols, type = "pred", terms = c("change_ei_conf", "biden_voter", "fttrump [0, 40, 80]"), ci.lvl = .95, title = "Predicted change in anger index", axis.title = c("Change in elec. confidence", "Change in anger index"), legend.title = "Biden voter")

p_bv_ft_ei_ols2 <- sjPlot::plot_model(model_biden_ei_ft_ols, type = "pred", terms = c("fttrump", "biden_voter", "change_ei_conf [-2, 0, 2]"), ci.lvl = .95, title = "Predicted change in anger index", axis.title = c("Trump thermometer", "Change in anger index"), legend.title = "Biden voter")

```

## Results: Panel survey

```{r anger-fttrump-mlm, fig.cap="Anger, Trump feeling thermometer rating, and elction confidence (MLM)", out.width="85%"}
p_ft_ei_mlm + theme_bw()
```

## Results: Panel survey


```{r ft-bv-ei, fig.cap="Three-way interaction of change in election confidence, Biden vote, and Trump rating (OLS).", out.width="85%"}
p_bv_ft_ei_ols2 + theme_bw()
```

## Discussion and conclusion

- Appears that legitimacy costs  primarily expressed through base mobilization (at least in polarized polities) 
- If the cost of a manipulated election is largely paid by angering a party's die-hard opponents, the cost may be relatively bearable for governments
  - Targeted repression + motivated reasoning and media control
 
- Democratic backsliding: the cost of rigging an election in a democracy is also lower in the base-mobilization model

## Discussion and conclusion

- These findings are pessimistic for democracy
- Help explain the numerous instances of failed electoral revolutions in recent history. 
- Base mobilization, anger, make false allegations appealing
- More research needed:
 - What conditions create legitimacy costs among moderates and winning-party supporters?
  - Parties' messaging strategies, the existence of credible non-partisan monitors, coalitional nature of parties?
  
## Thanks!

- Contact info:
 - cole.j.harvey@okstate.edu
 - www.colejharvey.com

## References {.allowframebreaks}
