---
title: "Appendix: Legitimacy costs in a 'rigged' election: Observational and survey-experimental evidence from the 2020 US election"
output:
    bookdown::pdf_document2: default
linkcolor: blue
bibliography: C:/Users/colej/Documents/Research projects/2020_election_survey/bib_ch.bib
#biblio-style: "apalike"
fontsize: 12pt
link-citations: yes
toc: true
csl: C:/Users/colej/Documents/Research projects/2020_election_survey/american-journal-of-political-science.csl
header-includes: \usepackage{booktabs}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(tidyverse)
library(lme4)
library(kableExtra)
library(modelsummary)
library(interactions)


anes_data <- read.csv(here::here("Data", "anes_modified_long.csv"))

survey.data <- read.csv(here::here("Data", "survey data.csv"))
survey.data <- survey.data %>% mutate(biden.winner = ifelse(vignette_winning_candidate == "biden", 1, 0))
survey.data <- survey.data %>% mutate(anger.index = (anger + outrage)/2)


anes_data <- anes_data %>% mutate(gender_factor = factor(profile_gender))
anes_data <- anes_data %>% mutate(race_factor = factor(profile_racethnicity))

## Dividing ANES respondents into Trump opponents, Trump moderates, and  Trump supporters
anes_data <- anes_data %>% mutate(trump_support_factor = ifelse(is.na(fttrump)==T, NA, ifelse(fttrump <=20, "opponent",
                                                                            ifelse(fttrump >=80, "supporter", ifelse(fttrump > 20 & fttrump <50, "lean_opponent", "lean_supporter")))))
anes_data <- anes_data %>% mutate(trump_support_factor = factor(trump_support_factor))

anes_data$trump_support_factor <- relevel(anes_data$trump_support_factor, ref = "opponent")

```




## Balance indicators

The figures presented below show balance statistics across a range of observables for the two treatment conditions in the survey experiment: winning candidate (Biden/Trump) and election integrity (fraudulent/clean). Figures \@ref(fig:loveplot-falsify-cont) and \@ref(fig:loveplot-falsify-cat) show the standardized means differences across the fraud treatment and control groups for continuous and categorical variables, respectively. They show that no covariate exceeds the rule-of-them imbalance threshold of 0.1. Likewise, Figure \@ref(fig:loveplot-winner-cat) shows no imbalance in the categorical covariates of the winning-candidate treatment. The exception is Figure \@ref(fig:loveplot-winner-cont), which does indicate some imbalance in the ideology and Trump rating of respondents in the Biden winning condition versus the Trump condition. Respondents randomized into the Biden-winning group were slightly more left-leaning on average, and slightly more disapproving of Trump, than those randomized into the Trump-winning group. Consequently, I control for ideology in the main models; *Trump rating* is already an explanatory variable included in the models.

```{r exp-summary}
survey.data <- survey.data %>% filter(vignette_laws == 0)

covariates <- survey.data %>% dplyr::select(biden_rating, biden.winner, vignette_control, vignette_falsify,  Socioeconomic.Status, ideology, interest_politics, votes_counted_fairly, election_officials_fair, voters_have_choice, trump_rating, Employment.Status, Employment.Status, Ethnicity, Highest.education.level.completed, Sex, age.x) %>% as_tibble()



covariates <- covariates %>% filter(age.x != "")
covariates <- covariates %>% filter(Employment.Status != "")
covariates <- covariates %>% filter(Employment.Status != "DATA EXPIRED")
covariates <- covariates %>% filter(Sex != "DATA EXPIRED")

covariates <- covariates %>% mutate(Employment.Status = factor(Employment.Status))
covariates <- covariates %>% mutate(biden_rating = as.numeric(biden_rating))
covariates <- covariates %>% mutate(trump_rating = as.numeric(trump_rating))
covariates <- covariates %>% mutate(Ethnicity = factor(Ethnicity))
covariates <- covariates %>% mutate(Highest.education.level.completed = factor(Highest.education.level.completed))
#covariates <- covariates %>% mutate(Nationality = factor(Nationality))
covariates <- covariates %>% mutate(Sex = factor(Sex))
covariates <- covariates %>% mutate(Socioeconomic.Status = as.numeric(Socioeconomic.Status))
#covariates <- covariates %>% mutate(U.S..Political.Affiliation = factor(U.S..Political.Affiliation))
covariates <- covariates %>% mutate(age.x = factor(age.x))

#datasummary_balance(~vignette_falsify, data=covariates)
#datasummary_balance(~biden.winner, data = covariates)



#covariates <- covariates %>% fastDummies::dummy_cols(select_columns = c("Employment.Status", "Ethnicity", "age.x", "Sex"))
```


```{r}

covariates_cont <- covariates %>% dplyr::select(-Employment.Status, -Ethnicity, -age.x, -Sex,
                                                -Highest.education.level.completed, -biden.winner)
covariates_cat <- covariates %>% dplyr::select(Employment.Status, Ethnicity, age.x, Sex,
                                                Highest.education.level.completed, biden.winner, vignette_falsify)
```

```{r loveplot-falsify-cont, fig.cap="Covariate balance plot, election integrity treatment (continuous variables)", warning=FALSE, message=FALSE}
balance_table_cont <- cobalt::bal.tab(covariates_cont, treat = "vignette_falsify", data = covariates_cont, stats = "m")
#treat <- covariates_cont$vignette_falsify
#col_w_smd(covariates_cont, treat=treat, na.rm = T)
cobalt::love.plot(balance_table_cont, stats = "mean.diffs", stars = "raw", thresholds = c(m = .1))
```



```{r loveplot-falsify-cat, fig.cap="Covariate balance plot, election integrity treatment (continuous variables)", warning=FALSE, message=FALSE}
balance_table_cat <- cobalt::bal.tab(covariates_cat, treat = "vignette_falsify", data = covariates_cat, stats = "m")
cobalt::love.plot(balance_table_cat, stats = "mean.diffs", stars = "raw", thresholds = c(m = .1))

```


```{r}
covariates_cont <- covariates %>% dplyr::select(-Employment.Status, -Ethnicity, -age.x, -Sex,
                                                -Highest.education.level.completed, -vignette_falsify)
covariates_cat <- covariates %>% dplyr::select(Employment.Status, Ethnicity, age.x, Sex,
                                                Highest.education.level.completed, biden.winner, vignette_falsify)
```

```{r loveplot-winner-cont, fig.cap="Covariate balance plot, winner treatment (continuous variables)", warning=FALSE, message=FALSE}

balance_table_cont <- cobalt::bal.tab(covariates_cont, treat = "biden.winner", data = covariates_cont, stats = "m")
#treat <- covariates_cont$vignette_falsify
#col_w_smd(covariates_cont, treat=treat, na.rm = T)
cobalt::love.plot(balance_table_cont, stats = "mean.diffs", stars = "raw", thresholds = c(m = .1))
```

```{r loveplot-winner-cat, fig.cap="Covariate balance plot, winner treatment (categorical variables)", warning=FALSE, message=FALSE}
balance_table_cat <- cobalt::bal.tab(covariates_cat, treat = "biden.winner", data = covariates_cat, stats = "m")
cobalt::love.plot(balance_table_cat, stats = "mean.diffs", stars = "raw", thresholds = c(m = .1))

```

\newpage

## Mediation analysis for anger and collective action

```{r experiment-protest-model}
model.demon.falsify <- lm(ca.index ~ trump_rating*vignette_falsify  + ideology, data = survey.data %>% filter(vignette_laws == 0 & biden.winner == 1))
#summary(model.demon.falsify)

p_exp_demon <- sjPlot::plot_model(model.demon.falsify, type = "pred", terms = c("trump_rating [0, 9]", "vignette_falsify"), title = "Predicted values of collective action index", axis.title = c("Trump rating", "CA index"), legend.title = "Fraud treatment")
```

The survey experiment also captured respondents' self-reported interest in participating in collective action following the election vignette they viewed. Respondents were asked how likely they would be to engage in six different forms of collective action: signing a petition, participating in a demonstration, donating to their party, donating to an organization that fights for fair elections, and doing something with fellow group members to "stop this." Responses were collected on a five-point scale; these responses were used to create an additive index for *collective action interest*. This variable is used in robustness checks reported here, both as a supplemental dependent variable and as the outcome variable for mediation analysis using anger as the intermediary.

Testing respondents' self-reported likelihood of engaging in collective action in response to the election comports with the findings on anger, though with even less support for the coalition-building hypothesis. In the control condition, strong Trump supporters report roughly three times the interest in collective action as Trump opponents, as Figure \@ref(fig:exp-demon-fig) shows. In the fraud condition, however, strong Trump opponents report only as much interest in challenging the election collectively as strong Trump supporters do in the control condition. This is noticeably different from Figure 1 in the main text, which shows Trump opponents expressing roughly the same level of anger as Trump supporters in the fraud condition. Thus, while we do observe some notional support for the coalition-expansion hypothesis, it is weaker among strong Trump opponents than what is found using anger as a dependent variable.

```{r exp-demon-fig, fig.cap="Trump support, election fraud, and protest (survey experiment)"}
p_exp_demon + theme_bw()
```

Mediation analysis shows that anger partially mediates this effect of fraud on collective action as predicted.[^5] Mediation analysis was performed using two models: one regressing *anger index* on the interaction of treatment status and Trump rating, and one regressing *collective action index* on *anger index* and the same interaction term. Both models control for ideology, as in the main models. The results in Table \@ref(tab:mediation-table) show that about forty-one percent of the total effect of the fraud treatment on collective-action intention is mediated by anger (shown here as the average conditional marginal effect). This shows that, while not the sole predictor of interest in collective action to redress election manipulation, anger is a powerful motivator.  

[^5]: The analysis was run using the `mediation` package in R (@Tingley2014).

```{r}
###Function to extract info from mediation summary
extract_mediation_summary <- function (x) { 

  clp <- 100 * x$conf.level
  isLinear.y <- ((class(x$model.y)[1] %in% c("lm", "rq")) || 
                   (inherits(x$model.y, "glm") && x$model.y$family$family == 
                      "gaussian" && x$model.y$family$link == "identity") || 
                   (inherits(x$model.y, "survreg") && x$model.y$dist == 
                      "gaussian"))

  printone <- !x$INT && isLinear.y

  if (printone) {

    smat <- c(x$d1, x$d1.ci, x$d1.p)
    smat <- rbind(smat, c(x$z0, x$z0.ci, x$z0.p))
    smat <- rbind(smat, c(x$tau.coef, x$tau.ci, x$tau.p))
    smat <- rbind(smat, c(x$n0, x$n0.ci, x$n0.p))

    rownames(smat) <- c("ACME", "ADE", "Total Effect", "Prop. Mediated")

  } else {
    smat <- c(x$d0, x$d0.ci, x$d0.p)
    smat <- rbind(smat, c(x$d1, x$d1.ci, x$d1.p))
    smat <- rbind(smat, c(x$z0, x$z0.ci, x$z0.p))
    smat <- rbind(smat, c(x$z1, x$z1.ci, x$z1.p))
    smat <- rbind(smat, c(x$tau.coef, x$tau.ci, x$tau.p))
    smat <- rbind(smat, c(x$n0, x$n0.ci, x$n0.p))
    smat <- rbind(smat, c(x$n1, x$n1.ci, x$n1.p))
    smat <- rbind(smat, c(x$d.avg, x$d.avg.ci, x$d.avg.p))
    smat <- rbind(smat, c(x$z.avg, x$z.avg.ci, x$z.avg.p))
    smat <- rbind(smat, c(x$n.avg, x$n.avg.ci, x$n.avg.p))

    rownames(smat) <- c("ACME (control)", "ACME (treated)", 
                        "ADE (control)", "ADE (treated)", "Total Effect", 
                        "Prop. Mediated (control)", "Prop. Mediated (treated)", 
                        "ACME (average)", "ADE (average)", "Prop. Mediated (average)")

  }

  colnames(smat) <- c("Estimate", paste(clp, "% CI Lower", sep = ""), 
                      paste(clp, "% CI Upper", sep = ""), "p-value")
  smat

}
```


```{r mediation-models}
mediation_model2 <- lm(anger.index ~ vignette_falsify*trump_rating + ideology, data = survey.data %>% filter(vignette_laws == 0 & biden.winner == 1 & is.na(ca.index)==F))
#summary(mediation_model2) #Step 2 passes: vignette increases anger

mediation_model3 <- lm(ca.index ~ anger.index + vignette_falsify*trump_rating + ideology, data = survey.data %>% filter(vignette_laws == 0 & biden.winner == 1 & is.na(ca.index)==F))

contcont <- mediation::mediate(mediation_model2, mediation_model3, sims=1000, treat="vignette_falsify", mediator="anger.index")

med_table <-extract_mediation_summary(summary(contcont))

for(i in 1:4){
 med_table[i,4] <- ifelse(med_table[i,4] == 0, "0***", 0)
}
med_table <- data.frame(med_table)
med_table$Estimate <- as.numeric(med_table$Estimate)
med_table$X95..CI.Lower <- as.numeric(med_table$X95..CI.Lower)
med_table$X95..CI.Upper <- as.numeric(med_table$X95..CI.Upper)

med_table <- med_table %>% rename(CI.Lower = X95..CI.Lower)
med_table <- med_table %>% rename(CI.Upper = X95..CI.Upper)



```

```{r mediation-table}
kable(med_table, digits = 3, booktabs = T, caption = "Mediated and total effect of fraud treatment on collective action") %>%
  kable_styling(position = "center")
```

Finally, as hinted by the above discussion, anger's mediating effect appears to increase in strength with Trump approval. In other words, Trump supporters' anger is more readily translated into collective action interest than Trump opponents'. Figure \@ref(fig:fig-ca-trump) shows the marginal effect of the fraud vignette on collective action interest by Trump support based on the outcome model of the mediation model (that is, controlling for anger). As the figure demonstrates, strong Trump opponents are not significantly more likely to express interest in collective action (though they do express high levels of anger). Meanwhile, the size of the marginal effect increases with Trump support. This is further supplementary evidence of the base-mobilization hypothesis. 

```{r}
p.ca.trump <- sjPlot::plot_model(mediation_model3, type = "pred", terms = c("trump_rating [0, 9]", "vignette_falsify"), title = "Predicted values of collective action index", axis.title = c("Trump rating", "CA index"), legend.title = "Fraud treatment")
```


```{r fig-ca-trump, fig.cap="Trump support, election fraud, and collective action"}
p.ca.trump + theme_bw()
```


## Trump as election winner in survey experiment

In the main text, results are presented from the treatment arm of the survey experiment in which Joe Biden was presented as the winner of the election; this aligns the experiment with the real-world outcome and enables comparison with the ANES data. The following figures show results from the same models for respondents in the other treatment arm, in which Donald Trump was presented as the winner. The results are symmetrical to those in the main text. Note that in these models, *Trump rating* has been replaced with *Biden rating*, a feeling thermometer rating of Joe Biden. Figure \@ref(fig:trump-win-anger) shows the same latent coalition-broadening effect of fraud on feelings of anger; in this case, respondents with low affinity for Joe Biden see the largest increases in anger from the control to fraud conditions---just as those with the lowest affinity for Donald Trump see the largest increases in the main models.

```{r trump-win-anger, fig.cap="Anger, Trump support, and fraud (Trump winner condition)"}

model.anger.falsify_twin <- lm(anger.index ~ biden_rating*vignette_falsify  + ideology, data = survey.data %>% filter(vignette_laws == 0 & biden.winner == 0))
#summary(model.anger.falsify_twin)

p_exp_anger <- sjPlot::plot_model(model.anger.falsify_twin, type = "pred", terms = c("biden_rating [0, 9]", "vignette_falsify"), title = "Predicted values of anger index", axis.title = c("Biden rating", "Anger index"),
                                  legend.title = "Fraud treatment")

p_exp_anger + theme_bw()
```

Likewise, Figure \@ref(fig:trump-win-ca) shows the same parallel marginal effects shown in Figure \@ref(fig:experiment-protest-model) in the main text. The effect is somewhat more modest than in the main models, but still, individuals with relatively moderate ratings of Joe Biden are significantly more interested in protest in the fraud condition than the strongest Biden supporters are in the control.

```{r trump-win-ca, fig.cap="Collective action, Trump support, and fraud (Trump winner condition)"}
model.demon.falsify.twin <- lm(ca.index ~ biden_rating*vignette_falsify  + ideology, data = survey.data %>% filter(vignette_laws == 0 & biden.winner == 0))
#summary(model.demon.falsify.twin)

p_exp_demon <- sjPlot::plot_model(model.demon.falsify.twin, type = "pred", terms = c("biden_rating [0, 9]", "vignette_falsify"), title = "Predicted values of collective action index", axis.title = c("Biden rating", "CA index"), legend.title = "Fraud treatment")

p_exp_demon + theme_bw()
```

Finally, mediation analysis shows that anger is an effective mediator between the fraud treatment and interest in post-election collective action for respondents who considered an election in which Donald Trump was the winner, just as it was for those who imagined Biden as the winner. The size of the mediation effect, however, is smaller for respondents in the Trump-winning group, as Table \@ref(tab:mediation-table-appendix) shows.  While anger mediated about 40 percent of the effect on collective-action interest in the Biden-wins group, only about 25 percent is mediated here. This suggests that Trump supporters are less willing to engage in post-election collective action against a fraudulent win than are Biden supporters (on average). This is further support for the base-mobilization hypothesis.

```{r mediation-models-app}
mediation_model_t2 <- lm(anger.index ~ vignette_falsify*biden_rating + ideology, data = survey.data %>% filter(vignette_laws == 0 & biden.winner == 0 & is.na(ca.index)==F))
#summary(mediation_model2) #Step 2 passes: vignette increases anger

mediation_model_t3 <- lm(ca.index ~ anger.index + vignette_falsify*biden_rating + ideology, data = survey.data %>% filter(vignette_laws == 0 & biden.winner == 0 & is.na(ca.index)==F))

contcont_t <- mediation::mediate(mediation_model_t2, mediation_model_t3, sims=1000, treat="vignette_falsify", mediator="anger.index")

med_table_t <-extract_mediation_summary(summary(contcont_t))

for(i in 1:4){
 med_table_t[i,4] <- ifelse(med_table_t[i,4] == 0, "0***", 0)
}
med_table_t <- data.frame(med_table_t)
med_table_t$Estimate <- as.numeric(med_table_t$Estimate)
med_table_t$X95..CI.Lower <- as.numeric(med_table_t$X95..CI.Lower)
med_table_t$X95..CI.Upper <- as.numeric(med_table_t$X95..CI.Upper)

med_table_t <- med_table_t %>% rename(CI.Lower = X95..CI.Lower)
med_table_t <- med_table_t %>% rename(CI.Upper = X95..CI.Upper)



```

```{r mediation-table-appendix}
kable(med_table_t, digits = 3, booktabs = T, caption = "Mediated and total effect of fraud treatment on collective action (Trump winning condition)") %>%
  kable_styling(position = "center")
```

```{r}
p.ca.biden <- sjPlot::plot_model(mediation_model_t3, type = "pred", terms = c("biden_rating [0, 9]", "vignette_falsify"), title = "Predicted values of collective action index", axis.title = c("Biden rating", "CA index"), legend.title = "Fraud treatment")
```


```{r fig-ca-biden, fig.cap="Biden support, election fraud, and collective action"}
p.ca.biden + theme_bw()
```

## Feeling thermometer instead of categorical Trump support

For the models in the main text, ANES respondents were grouped into four categories based on their evaluation of President Trump on a 100-point feeling thermometer. Here, the raw feeling thermometer scores are used as a robustness check. Figure \@ref(fig:anger-fttrump-mlm) and Table \@ref(tab:fttrump-table) shows the results from a model equivalent to Model 2 in the main text. Figure \@ref(fig:anger-fttrump-mlm), in particular, shows how reported feelings of anger shift across the pre- and post-election period according to Trump support and confidence in election integrity. In line with the categorical results in the main text, strong Trump supporters report roughly equal, high levels of anger before and after the election when they have low confidence in elections (shown in the left-most panel in the figure). When they report high confidence in the election (the rightmost panel), they report significantly less anger after the election. Political moderates and Trump opponents generally report less anger after the election across all three levels of election confidence; this provides further support for the base-mobilization hypothesis.


```{r anger_models_mlm}
anes_data <- anes_data %>% mutate(period = factor(period))

model_ft_ei_mlm <- lmer(anger_index ~ period*fttrump*ei_index  + trump_voter +  pid7x +
                     gender_factor + race_factor + 
                     profile_educ5 +
                     profile_income + profile_newskeepup + covid_fin + covid_worry + persfin + (1 | caseid)
                    , data = anes_data)

  #summary(model_ft_ei_mlm)
p_ft_ei_mlm <- sjPlot::plot_model(model_ft_ei_mlm, type = "pred", terms = c("fttrump", "period", "ei_index [1, 3, 5]"), title = "Predicted values of anger index", axis.title = c("Trump rating", "Anger index"),
                                  legend.title = "Post-election period")



model_biden_ei_mlm <- lmer(anger_index ~ period*trump_voter*ei_index  + fttrump +  pid7x +
                     gender_factor + race_factor + 
                     profile_educ5 +
                     profile_income + profile_newskeepup + covid_fin + covid_worry + persfin + (1 | caseid)
                    , data = anes_data)

  #summary(model_mlm)
p_bv_ei_mlm <- sjPlot::plot_model(model_biden_ei_mlm, type = "pred", terms = c("period", "ei_index [1, 3, 5]", "trump_voter"), title = "Predicted values of anger index", axis.title = c("Post-election period", "Anger index"),
                                  legend.title = "Elec. confidence")

```

```{r fttrump-table, warning=FALSE}
options(modelsummary_html = 'kableExtra')
library(kableExtra)

mods <- list()

mods[["2"]] <- model_ft_ei_mlm



##Renaming the variables for the table


modelsummary(mods, title = "ANES multilevel models", longtable = T, booktabs = T, stars = T, 
             coef_rename=c("period1" = "Post-election",
                           'fttrump' = 'Trump rating',
                           'trump_voter' = 'Trump voter',
                            'ei_index'= 'Elec. confidence',
                            'pid7x'   =  'Party ID',
                           "persfin" = "Financial worry",
                            'gender_factor2' = 'Female',
                            'race_factor2' = 'Black',
                           'race_factor3' = 'Other, non-Hispanic',
                           'race_factor4' = 'Hispanic',
                           'profile_educ5'  =  'Education level',
                           'profile_income' = 'Income',
                           'profile_newskeepup'   = 'Keep up with news',
                           'covid_fin'  = 'Covid financial concern',
                           'covid_worry' = 'Covid health worry',
                           "period1:fttrump" = "Post-election x Trump rating",
                           "period1:ei_index" = "Post-election x Elec. confidence",
                           "fttrump:ei_index" = "Trump rating x Elec. confidence",
                           "period1:fttrump:ei_index" = "Post-election x Trump rating x Elec. conf."),
             gof_omit = 'Cond.|Marg.|F|Log.') %>%
 # footnote(general = "* = p < .1, ** = p < .05, *** = p <.01",threeparttable = T) %>%
  kable_styling(latex_options =c("repeat_header"))
```


```{r anger-fttrump-mlm, fig.cap="Anger, Trump support, and election confidence by pre-election (0) and post-election (1) panel wave"}
p_ft_ei_mlm + theme_bw()
```

\newpage

## References
